---
title: "E13.5 cortex"
output:
    html_notebook:
      css: html-md-01.css
      fig_caption: yes
      highlight: haddock
      number_sections: yes
      theme: spacelab
      toc: yes
      toc_float: true
      collapsed: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE,
  tidy = TRUE, tidy.opts=list(width.cutoff=40)
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(COTAN)
library(data.table)
library(Matrix)
library(ggrepel)
#library(latex2exp)
```

```{r}
mycolours <- c("A" = "#8491B4B2","B"="#E64B35FF")
my_theme = theme(axis.text.x = element_text(size = 14, angle = 0, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF" ),
                 axis.text.y = element_text( size = 14, angle = 0, hjust = 0, vjust = .5, face = "plain", colour ="#3C5488FF"),
                 axis.title.x = element_text( size = 14, angle = 0, hjust = .5, vjust = 0, face = "plain", colour ="#3C5488FF"),
                 axis.title.y = element_text( size = 14, angle = 90, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF"))
```


```{r}
data_dir = "Data/"
```
```{r}
data = as.data.frame(fread(paste(data_dir,"GSM2861511_E135_Only_Cortical_Cells_DGE.txt.gz", sep = "/")))
data = as.data.frame(data)
rownames(data) = data$V1
data = data[,2:ncol(data)]
data[1:10,1:10]
```

Define a directory where the ouput will be stored.
```{r}
out_dir = "Data/"
```

# Analitical pipeline

Inizialise the COTAN object with the row count table and the metadata for the experiment.

```{r}
obj = new("scCOTAN",raw = data)
obj = initRaw(obj,GEO="GSM2861511" ,sc.method="DropSeq",cond = "mouse cortex E13.5")
```

Now we can start the cleaning. Analysis requires and starts from a matrix of raw UMI counts after removing possible cell doublets or multiplets and low quality or dying cells (with too high mtRNA percentage, easily done with Seurat or other tools).

If we do not want to consider the mithocondrial genes we can remove them before starting the analysis.
```{r}
genes_to_rem = rownames(obj@raw[grep('^mt', rownames(obj@raw)),]) #genes to remove : mithocondrial
obj@raw = obj@raw[!rownames(obj@raw) %in% genes_to_rem,]
cells_to_rem = colnames(obj@raw[which(colSums(obj@raw) == 0)])
obj@raw = obj@raw[,!colnames(obj@raw) %in% cells_to_rem]
```
We want also to define a prefix to identify the sample.

```{r}
t = "E13_cortex_clean"

print(paste("Condition ",t,sep = ""))
#--------------------------------------
n_cells = length(colnames(obj@raw))
print(paste("n cells", n_cells, sep = " "))

n_it = 1

#obj@raw = obj@raw[rownames(cells), colnames(cells)]
```
# Data cleaning
First we create the directory to store all information regardin the data cleaning.
```{r}
if(!file.exists(out_dir)){
  dir.create(file.path(out_dir))
}

if(!file.exists(paste(out_dir,"cleaning", sep = ""))){   
  dir.create(file.path(out_dir, "cleaning"))
}
```

```{r}
ttm = clean(obj)

obj = ttm$object

ttm$pca.cell.2
```

Run this when B cells need to be removed.
```{r}
pdf(paste(out_dir,"cleaning/",t,"_",n_it,"_plots_before_cells_exlusion.pdf", sep = ""))
ttm$pca.cell.2
ggplot(ttm$D, aes(x=n,y=means)) + geom_point() +
  geom_text_repel(data=subset(ttm$D, n > (max(ttm$D$n)- 15) ), aes(n,means,label=rownames(ttm$D[ttm$D$n > (max(ttm$D$n)- 15),])),
                  nudge_y      = 0.05,
                  nudge_x      = 0.05,
                  direction    = "x",
                  angle        = 90,
                  vjust        = 0,
                  segment.size = 0.2)+
  ggtitle("B cell group genes mean expression")+my_theme +
  theme(plot.title = element_text(color = "#3C5488FF", size = 20, face = "italic",vjust = - 5,hjust = 0.02 ))
dev.off()

if (length(ttm$cl1) < length(ttm$cl2)) {
  to_rem = ttm$cl1
}else{
  to_rem = ttm$cl2
}
n_it = n_it+1
obj@raw = obj@raw[,!colnames(obj@raw) %in% to_rem]
#obj@raw = obj@raw[rownames(obj@raw) %in% rownames(cells),colnames(obj@raw) %in% colnames(cells)]
gc()

ttm = clean(obj)
#ttm = clean.sqrt(obj, cells)
obj = ttm$object

ttm$pca.cell.2

```
```{r echo=FALSE}
pdf(paste(out_dir,"cleaning/",t,"_",n_it,"_plots_before_cells_exlusion.pdf", sep = ""))
ttm$pca.cell.2
ggplot(ttm$D, aes(x=n,y=means)) + geom_point() +
  geom_text_repel(data=subset(ttm$D, n > (max(ttm$D$n)- 15) ), aes(n,means,label=rownames(ttm$D[ttm$D$n > (max(ttm$D$n)- 15),])),
                  nudge_y      = 0.05,
                  nudge_x      = 0.05,
                  direction    = "x",
                  angle        = 90,
                  vjust        = 0,
                  segment.size = 0.2)+
  ggtitle("B cell group genes mean expression")+my_theme +
  theme(plot.title = element_text(color = "#3C5488FF", size = 20, face = "italic",vjust = - 5,hjust = 0.02 ))
dev.off()

if (length(ttm$cl1) < length(ttm$cl2)) {
  to_rem = ttm$cl1
}else{
  to_rem = ttm$cl2
}
n_it = n_it+1
obj@raw = obj@raw[,!colnames(obj@raw) %in% to_rem]
#obj@raw = obj@raw[rownames(obj@raw) %in% rownames(cells),colnames(obj@raw) %in% colnames(cells)]
gc()

ttm = clean(obj)
#ttm = clean.sqrt(obj, cells)
obj = ttm$object

ttm$pca.cell.2

```

Run this only in the last iteration, instead the previus code, when B cells group has not to be removed
```{r}
pdf(paste(out_dir,"cleaning/",t,"_",n_it,"_plots_before_cells_exlusion.pdf", sep = ""))
ttm$pca.cell.2
ggplot(ttm$D, aes(x=n,y=means)) + geom_point() +
  geom_text_repel(data=subset(ttm$D, n > (max(ttm$D$n)- 15) ), aes(n,means,label=rownames(ttm$D[ttm$D$n > (max(ttm$D$n)- 15),])),
                  nudge_y      = 0.05,
                  nudge_x      = 0.05,
                  direction    = "x",
                  angle        = 90,
                  vjust        = 0,
                  segment.size = 0.2)+
  ggtitle(label = "B cell group genes mean expression", subtitle = " - B group NOT removed -")+my_theme +
  theme(plot.title = element_text(color = "#3C5488FF", size = 20, face = "italic",vjust = - 10,hjust = 0.02 ),
        plot.subtitle = element_text(color = "darkred",vjust = - 15,hjust = 0.01 ))

dev.off()
```
To color the pca based on nu_j (so the cells' efficiency)

```{r}
nu_est = round(obj@nu, digits = 7)

plot.nu <-ggplot(ttm$pca_cells,aes(x=PC1,y=PC2, colour = log(nu_est)))

plot.nu = plot.nu + geom_point(size = 1,alpha= 0.8)+
  scale_color_gradient2(low = "#E64B35B2",mid =  "#4DBBD5B2", high =  "#3C5488B2" ,
                        midpoint = log(mean(nu_est)),name = "ln (nu)")+
  ggtitle("Cells PCA coloured by cells efficiency") +
  my_theme +  theme(plot.title = element_text(color = "#3C5488FF", size = 20),
                    legend.title=element_text(color = "#3C5488FF", size = 14,face = "italic"),
                    legend.text = element_text(color = "#3C5488FF", size = 13),
                    legend.key.width = unit(2, "mm"),
                    legend.position="right")

pdf(paste(out_dir,"cleaning/",t,"_plots_PCA_efficiency_colored.pdf", sep = ""))
plot.nu
dev.off()

plot.nu
```
The next part is use to remove the cells with efficiency too low.
```{r}
nu_df = data.frame("nu"= sort(obj@nu), "n"=c(1:length(obj@nu)))

ggplot(nu_df, aes(x = n, y=nu)) + 
    geom_point(colour = "#8491B4B2", size=1)+
    my_theme #+ ylim(0,1) + xlim(0,70)

```

We can zoom on the smalest values and, if we detect a clear elbow, we can decide to remove the cells.
```{r}
yset = 0.18#threshold to remove low UDE cells
plot.ude <- ggplot(nu_df, aes(x = n, y=nu)) + 
    geom_point(colour = "#8491B4B2", size=1) + 
    my_theme + ylim(0.,0.5) + xlim(0,200) +
    geom_hline(yintercept=yset, linetype="dashed", color = "darkred") +
    annotate(geom="text", x=500, y=0.5, 
             label=paste("to remove cells with nu < ",yset,sep = " "), 
             color="darkred", size=4.5)


pdf(paste(out_dir,"cleaning/",t,"_plots_efficiency.pdf", sep = ""))
plot.ude
dev.off()

plot.ude
```
We also save the defined treshold in the metadata and re run the estimation
```{r}
obj@meta[(nrow(obj@meta)+1),1:2] = c("Threshold low UDE cells:",yset)

to_rem = rownames(nu_df[which(nu_df$nu < yset),])

obj@raw = obj@raw[, !colnames(obj@raw) %in% to_rem]
```

Repeat the estimation after the cells are removed

```{r}
ttm = clean(obj)
obj = ttm$object
ttm$pca.cell.2
```


Just to check again, we plot the final efficiency colored PCA

```{r}
nu_est = round(obj@nu, digits = 7)
plot.nu <-ggplot(ttm$pca_cells,aes(x=PC1,y=PC2, colour = log(nu_est)))
plot.nu = plot.nu + geom_point(size = 2,alpha= 0.8)+
  scale_color_gradient2(low = "#E64B35B2",mid =  "#4DBBD5B2", high =  "#3C5488B2" ,
                        midpoint = log(mean(nu_est)),name = "ln(nu)")+
  ggtitle("Cells PCA coloured by cells efficiency: last") +
  my_theme +  theme(plot.title = element_text(color = "#3C5488FF", size = 20),
                    legend.title=element_text(color = "#3C5488FF", size = 14,face = "italic"),
                    legend.text = element_text(color = "#3C5488FF", size = 13),
                    legend.key.width = unit(2, "mm"),
                    legend.position="right")

pdf(paste(out_dir,"cleaning/",t,"_plots_PCA_efficiency_colored_FINAL.pdf", sep = ""))
plot.nu
dev.off()

plot.nu
```
COTAN analysis: in this part all the contingency tables are computed and used to get the statistics (S)
To storage efficincy of all the observed tables only the yes_yes is stored. 
Note that if will be necessary re-coputing the yes-yes table, the yes-yes table should be cancelled before running cotan_analysis.

```{r eval=FALSE, include=TRUE}
# obj@yes_yes = c()
obj = cotan_analysis(obj)
```

COEX evaluation and storing 
```{r eval=FALSE, include=TRUE}
saveRDS(obj,file = paste(out_dir,t,".cotan.RDS", sep = ""))

obj = get.coex(obj)

# saving the structure 
saveRDS(obj,file = paste(out_dir,t,".cotan.RDS", sep = ""))
```

Automatic data elaboration (without cleaning)
```{r}
obj2 = automatic.COTAN.object.creation(df = data,GEO="GSM2861510" ,sc.method="Drop-seq",cond = "E13_cortex",
                                       out_dir =  "Data/")
```

# Analysis of the elaborated data

In this case the test possible macroscopic differences between cleaned and not cleaned data.

The initial cell number was:
```{r}
obj = readRDS(paste(out_dir,t,".cotan.RDS", sep = ""))
obj@meta
```

While the final number of cells was:
```{r}
ncol(obj@raw)
```

So were removed:
```{r}
as.numeric(obj@meta[3,2]) - ncol(obj@raw)
```

cells


## GDI with cleaning

```{r echo=TRUE}
quant.p = get.GDI(obj)

head(quant.p)
```
In the third column of this dataframe is reported the percentage of cells expressing the gene.

```{r echo=TRUE}
NPGs=c("Nes","Vim","Sox2","Sox1","Notch1", "Hes1","Hes5","Pax6") #,"Hes3"
PNGs=c("Map2","Tubb3","Neurod1","Nefm","Nefl","Dcx","Tbr1")
hk=c("Calm1","Cox6b1","Ppia","Rpl18","Cox7c","Erh","H3f3a","Taf1","Taf2","Gapdh","Actb","Golph3", "Mtmr12", "Zfr", "Sub1", "Tars", "Amacr")

text.size = 12

quant.p$highlight = with(quant.p, ifelse(rownames(quant.p) %in% NPGs, "NPGs",
                                                 ifelse(rownames(quant.p) %in% hk,"Constitutive" ,
                                                               ifelse(rownames(quant.p) %in% PNGs,"PNGs" , "normal"))))

textdf <- quant.p[rownames(quant.p) %in% c(NPGs,hk,PNGs), ]
mycolours <- c("Constitutive" = "#00A087FF","NPGs"="#E64B35FF","PNGs"="#F39B7FFF","normal" = "#8491B4B2")
f1 = ggplot(subset(quant.p,highlight == "normal" ), aes(x=sum.raw.norm, y=GDI)) +  geom_point(alpha = 0.1, color = "#8491B4B2", size=2.5)
GDI_plot = f1 +  geom_point(data = subset(quant.p,highlight != "normal"  ), aes(x=sum.raw.norm, y=GDI, colour=highlight),size=2.5,alpha = 0.8) +
  geom_hline(yintercept=quantile(quant.p$GDI)[4], linetype="dashed", color = "darkblue") +
  geom_hline(yintercept=quantile(quant.p$GDI)[3], linetype="dashed", color = "darkblue") +
  geom_hline(yintercept=1.5, linetype="dotted", color = "red", size= 0.5) +
  scale_color_manual("Status", values = mycolours)  +
  scale_fill_manual("Status", values = mycolours)  +
  xlab("log normalized counts")+ylab("GDI")+
  geom_label_repel(data =textdf , aes(x=sum.raw.norm, y=GDI, label = rownames(textdf),fill=highlight),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction ="both",
                   na.rm=TRUE,
                   seed = 1234) +
  geom_label_repel(data =textdf , aes(x=sum.raw.norm, y=GDI, label = rownames(textdf)),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 0.8, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234) +
  theme(axis.text.x = element_text(size = text.size, angle = 0, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF" ),
        axis.text.y = element_text( size = text.size, angle = 0, hjust = 0, vjust = .5, face = "plain", colour ="#3C5488FF"),  
        axis.title.x = element_text( size = text.size, angle = 0, hjust = .5, vjust = 0, face = "plain", colour ="#3C5488FF"),
        axis.title.y = element_text( size = text.size, angle = 90, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF"),
        legend.title = element_blank(),
        legend.text = element_text(color = "#3C5488FF",face ="italic" ),
        legend.position = "bottom")  # titl)
legend <- cowplot::get_legend(GDI_plot)
GDI_plot =GDI_plot + theme(
        legend.position = "none") 
GDI_plot

```

## GDI without cleaning

```{r echo=TRUE}
quant.p = get.GDI(obj2)

head(quant.p)
```
In the third column of this dataframe is reported the percentage of cells expressing the gene.

```{r echo=TRUE}
NPGs=c("Nes","Vim","Sox2","Sox1","Notch1", "Hes1","Hes5","Pax6") #,"Hes3"
PNGs=c("Map2","Tubb3","Neurod1","Nefm","Nefl","Dcx","Tbr1")
hk=c("Calm1","Cox6b1","Ppia","Rpl18","Cox7c","Erh","H3f3a","Taf1","Taf2","Gapdh","Actb","Golph3", "Mtmr12", "Zfr", "Sub1", "Tars", "Amacr")

text.size = 12

quant.p$highlight = with(quant.p, ifelse(rownames(quant.p) %in% NPGs, "NPGs",
                                                 ifelse(rownames(quant.p) %in% hk,"Constitutive" ,
                                                               ifelse(rownames(quant.p) %in% PNGs,"PNGs" , "normal"))))

textdf <- quant.p[rownames(quant.p) %in% c(NPGs,hk,PNGs), ]
mycolours <- c("Constitutive" = "#00A087FF","NPGs"="#E64B35FF","PNGs"="#F39B7FFF","normal" = "#8491B4B2")
f1 = ggplot(subset(quant.p,highlight == "normal" ), aes(x=sum.raw.norm, y=GDI)) +  geom_point(alpha = 0.1, color = "#8491B4B2", size=2.5)
GDI_plot = f1 +  geom_point(data = subset(quant.p,highlight != "normal"  ), aes(x=sum.raw.norm, y=GDI, colour=highlight),size=2.5,alpha = 0.8) +
  geom_hline(yintercept=quantile(quant.p$GDI)[4], linetype="dashed", color = "darkblue") +
  geom_hline(yintercept=quantile(quant.p$GDI)[3], linetype="dashed", color = "darkblue") +
  geom_hline(yintercept=1.5, linetype="dotted", color = "red", size= 0.5) +
  scale_color_manual("Status", values = mycolours)  +
  scale_fill_manual("Status", values = mycolours)  +
  xlab("log normalized counts")+ylab("GDI")+
  geom_label_repel(data =textdf , aes(x=sum.raw.norm, y=GDI, label = rownames(textdf),fill=highlight),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction ="both",
                   na.rm=TRUE,
                   seed = 1234) +
  geom_label_repel(data =textdf , aes(x=sum.raw.norm, y=GDI, label = rownames(textdf)),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 0.8, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234) +
  theme(axis.text.x = element_text(size = text.size, angle = 0, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF" ),
        axis.text.y = element_text( size = text.size, angle = 0, hjust = 0, vjust = .5, face = "plain", colour ="#3C5488FF"),  
        axis.title.x = element_text( size = text.size, angle = 0, hjust = .5, vjust = 0, face = "plain", colour ="#3C5488FF"),
        axis.title.y = element_text( size = text.size, angle = 90, hjust = .5, vjust = .5, face = "plain", colour ="#3C5488FF"),
        legend.title = element_blank(),
        legend.text = element_text(color = "#3C5488FF",face ="italic" ),
        legend.position = "bottom")  # titl)
legend <- cowplot::get_legend(GDI_plot)
GDI_plot =GDI_plot + theme(
        legend.position = "none") 
GDI_plot

```

## Heatmaps with cleaning

For the Gene Pair Analysis, we can plot an heatmap with the coex values between two genes sets.
To do so we need to define, in a list, the different gene sets (list.genes). Then in the function parameter sets we can decide which sets need to be considered (in the example from 1 to 3). In the condition parameter we should insert an array with each file name prefix to be considered (in the example, the file names is "E17.5_cortex").


```{r}
list.genes = list("Ref.col"= PNGs, "NPGs"=NPGs, "Const."=hk )

plot_heatmap(df_genes = list.genes,sets = c(1:3),conditions = "E13_cortex_clean",dir = "Data/")
```
## Heatmaps without cleaning
```{r}
list.genes = list("Ref.col"= PNGs, "NPGs"=NPGs, "Const."=hk )

plot_heatmap(df_genes = list.genes,sets = c(1:3),conditions = "E13_cortex",dir = "Data/")
```

```{r}
print(sessionInfo())
```


