---
title: "Gene_clustering"
output:
  html_document:
    collapsed: no
    css: html-md-01.css
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
  html_notebook:
    collapsed: no
    css: html-md-01.css
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 7
)
```

```{r setup}

library(factoextra)
library(COTAN)
library(ggrepel)
library(Rtsne)
library(plotly)
library(tidyverse)
library(htmlwidgets)
library(MASS)
library(dendextend)
library(grid)
library(ggpubr)
```

To demostrate how to do the gene clustering usign COTAN we begin importing the COTAN object that stores all elaborated data and, in this case, regarding a mouse embrionic cortex dataset (developmental stage E17.5).

```{r}
input_dir = "Data/"
layers = list("L1"=c("Reln","Lhx5"), "L2/3"=c("Satb2","Cux1"), "L4"=c("Rorb","Sox5") , "L5/6"=c("Bcl11b","Fezf2") , "Prog"=c("Vim","Hes1"))
#objE17 = readRDS(file = paste(input_dir,"E17.5_cortex.cotan.RDS", sep = ""))
objE17 = readRDS(file = paste(input_dir,"E17_cortex_cl2.cotan.RDS", sep = ""))
```

```{r}
g.space = get.gene.coexpression.space(objE17, 
                                      n.genes.for.marker = 25,
                                      primary.markers = unlist(layers))
```
```{r}
g.space = as.data.frame(as.matrix(g.space))

coex.pca.genes <- prcomp(t(g.space),
                 center = TRUE,
                 scale. = F) 

fviz_eig(coex.pca.genes, addlabels=TRUE,ncp = 10)
#fviz_eig(coex.pca.genes, choice = "eigenvalue", addlabels=TRUE)
```
Hierarchical clustering

```{r include=FALSE}
hc.norm = hclust(dist(g.space), method = "ward.D2")

dend <- as.dendrogram(hc.norm)

pca_1 = as.data.frame(coex.pca.genes$rotation[,1:10])
pca_1 = pca_1[order.dendrogram(dend),]

cut = cutree(hc.norm, k = 7, order_clusters_as_data = F)

#- Next lines are only to color and plot the secondary markers

tmp = get.pval(object = objE17,gene.set.col =unlist(layers),gene.set.row = colnames(g.space))
for (m in unlist(layers)) {
  tmp = as.data.frame(tmp[order(tmp[,m]),])
  tmp$rank = c(1:nrow(tmp))
  colnames(tmp)[ncol(tmp)] = paste("rank",m,sep = ".")
  }
rank.genes = tmp[,(length(unlist(layers))+1):ncol(tmp)]
for (c in c(1:length(colnames(rank.genes)))) {
  colnames(rank.genes)[c] =strsplit(colnames(rank.genes)[c], split='.',fixed = T)[[1]][2]
}

L1 = rowSums(rank.genes[,layers[[1]]])
L1[layers[[1]]] = 1
L2 = rowSums(rank.genes[,layers[[2]]])
L2[layers[[2]]] = 1
L4 = rowSums(rank.genes[,layers[[3]]])
L4[layers[[3]]] = 1
L5 =rowSums(rank.genes[,layers[[4]]])
L5[layers[[4]]] = 1
P = rowSums(rank.genes[,layers[[5]]])
P[layers[[5]]] = 1
col.secondary = merge(L1,L2,by="row.names",all.x=TRUE)
colnames(col.secondary)[2:3] = c("L1","L2")
rownames(col.secondary) = col.secondary$Row.names
col.secondary = col.secondary[,2:ncol(col.secondary)]
col.secondary = merge(col.secondary,L4,by="row.names",all.x=TRUE)
colnames(col.secondary)[ncol(col.secondary)] = "L4"
rownames(col.secondary) = col.secondary$Row.names
col.secondary = col.secondary[,2:ncol(col.secondary)]
col.secondary = merge(col.secondary,L5,by="row.names",all.x=TRUE)
colnames(col.secondary)[ncol(col.secondary)] = "L5"
rownames(col.secondary) = col.secondary$Row.names
col.secondary = col.secondary[,2:ncol(col.secondary)]
col.secondary = merge(col.secondary,P,by="row.names",all.x=TRUE)
colnames(col.secondary)[ncol(col.secondary)] = "P"
rownames(col.secondary) = col.secondary$Row.names
col.secondary = col.secondary[,2:ncol(col.secondary)]

#  this part is to check that we will color as secondary markers only the genes linked to the
# primary with positive coex
temp.coex = as.matrix(objE17@coex[rownames(objE17@coex) %in% rownames(col.secondary),
                          colnames(objE17@coex) %in% unlist(layers)])
for (n in rownames(col.secondary)) {
  if(any(temp.coex[n,c("Reln","Lhx5")] < 0)){
    col.secondary[n,"L1"] = 100000
  }
  if(any(temp.coex[n,c("Cux1","Satb2")] < 0)){
    col.secondary[n,"L2"] = 100000
  }
  if(any(temp.coex[n,c("Rorb","Sox5")] < 0)){
    col.secondary[n,"L4"] = 100000
  }
  if(any(temp.coex[n,c("Bcl11b","Fezf2")] < 0)){
    col.secondary[n,"L5"] = 100000
  }
  if(any(temp.coex[n,c("Vim","Hes1")] < 0)){
    col.secondary[n,"P"] = 100000
  }
}

mylist.names <- c("L1", "L2", "L4","L5","P")
pos.link  <- vector("list", length(mylist.names))
names(pos.link) <- mylist.names
for (g in rownames(col.secondary)) {
  if(length( which(col.secondary[g,] == min(col.secondary[g,]))) == 1 ){
  pos.link[[which(col.secondary[g,] == min(col.secondary[g,])) ]] = 
    c(pos.link[[which(col.secondary[g,] == min(col.secondary[g,])) ]], g)
  }
}
# ----


pca_1$highlight = with(pca_1, 
          ifelse(rownames(pca_1) %in% pos.link$L5, "genes related to L5/6",
          ifelse(rownames(pca_1) %in% pos.link$L2 , "genes related to L2/3",
          ifelse(rownames(pca_1) %in% pos.link$P , "genes related to Prog" ,
          ifelse(rownames(pca_1) %in% pos.link$L1 , "genes related to L1" ,
          ifelse(rownames(pca_1) %in% pos.link$L4 ,"genes related to L4" ,
      "not marked"))))))

# But sort them based on their order in dend:
#colors_to_use <- pca_1$highlight[order.dendrogram(dend)]

#mycolours <- c("genes related to L5/6" = "#3C5488FF","genes related to L2/3"="#F39B7FFF","genes related to Prog"="#4DBBD5FF","genes related to L1"="#E64B35FF","genes related to L4" = "#91D1C2FF", "not marked"="#B09C85FF")
pca_1$hclust = cut

pca_1$colors = NA
pca_1[pca_1$highlight == "genes related to L5/6", "colors"] = "#3C5488FF"
pca_1[pca_1$highlight == "genes related to L2/3","colors"] = "#F39B7FFF"
pca_1[pca_1$highlight == "genes related to Prog","colors"] = "#4DBBD5FF"
pca_1[pca_1$highlight == "genes related to L1","colors"] = "#E64B35FF"
pca_1[pca_1$highlight == "genes related to L4","colors"] = "#91D1C2FF"
pca_1[pca_1$highlight == "not marked","colors"] = "#B09C85FF"




dend =branches_color(dend,k=7,col=c("#4DBBD5FF","#91D1C2FF","#E64B35FF","gray80","#3C5488FF","#F39B7FFF","gray80" ),groupLabels = T)
dend =color_labels(dend,k=7,labels = rownames(pca_1),col=pca_1$colors)


dend %>%
  dendextend::set("labels", ifelse(labels(dend) %in% rownames(pca_1)[rownames(pca_1) %in% colnames(g.space)] ,labels(dend),"")) %>%
  #  set("branches_k_color", value = c("gray80","#4DBBD5FF","#91D1C2FF" ,"gray80","#F39B7FFF","#E64B35FF","#3C5488FF"), k = 7) %>%
 plot(horiz=F, axes=T,ylim = c(0,80))
```
```{r include=FALSE}
cluster = cut
cluster[cluster == 1] = "#4DBBD5FF"
cluster[cluster == 2] = "#91D1C2FF"
cluster[cluster == 3] =  "#E64B35FF"
cluster[cluster == 4] = "#B09C85FF"
cluster[cluster == 5] = "#3C5488FF"
cluster[cluster == 6] = "#F39B7FFF"
cluster[cluster == 7] = "#B09C85FF"


plot.new()
plot(dend,horiz=T, axes=T,xlim = c(100,0),leaflab = "none")
abline(v = 47, lty = 2)
colored_bars(cluster,dend,horiz = T,sort_by_labels_order = F,y_shift = 1,
               rowLabels= "" )
gridGraphics::grid.echo()
tree <- grid.grab()
```
```{r fig.height=10, fig.width=7}
par_pca = pca_1[colnames(g.space)[colnames(g.space) %in% rownames(pca_1)],]
#plot N 1
p1 <- ggparagraph(text = paste0(rownames(par_pca[par_pca$hclust == par_pca["Fezf2","hclust"],]), collapse = ", "), 
                  face = "italic", 
                  size =10, 
                  color = "#3C5488FF")

#plot N 2
p2 = ggparagraph(text = paste0(rownames(par_pca[par_pca$hclust == unique(par_pca$hclust)[!unique(par_pca$hclust) %in% unique(par_pca[unlist(layers),"hclust"])][1],]), collapse = ", "), 
                 face = "italic", 
                 size =10, 
                 color = "gray")

#plot N 3
p3 = ggparagraph(text = paste0(rownames(par_pca[par_pca$hclust == unique(par_pca$hclust)[!unique(par_pca$hclust) %in% unique(par_pca[unlist(layers),"hclust"])][2],]), collapse = ", "), 
                 face = "italic", 
                 size =10, 
                 color = "gray")
#plot N 4
p4 = ggparagraph(text = paste0(rownames(par_pca[par_pca$hclust == par_pca["Reln","hclust"],]), collapse = ", "), 
                 face = "italic", 
                 size =10, 
                 color = "#E64B35FF")

#plot N 5
p5 = ggparagraph(text = paste0(rownames(par_pca[par_pca$hclust == par_pca["Cux1","hclust"],]), collapse = ", "), 
                 face = "italic", 
                 size =10, 
                 color = par_pca["Cux1","colors"])
#plot N 6
p6 = ggparagraph(text = paste0(rownames(par_pca[par_pca$hclust == par_pca["Rorb","hclust"],]), collapse = ", "), 
                 face = "italic", 
                 size =10, 
                 color = par_pca["Rorb","colors"])
#plot N 7
p7 = ggparagraph(text = paste0(rownames(par_pca[par_pca$hclust == par_pca["Vim","hclust"],]), collapse = ", "), 
                 face = "italic", 
                 size =10, 
                 color = par_pca["Vim","colors"])

w = ggparagraph(text = " ", 
                 face = "italic", 
                 size =10, 
                 color = "white")

pp =ggarrange(p3,p5,p1,p2,p4,p6,p7,w,
          ncol = 1, nrow = 8,
          heights = c(0.1,0.15,0.23, 0.1, 0.2, 0.2, 0.1, 0.35))


    
lay <- rbind(c(1,NA),
             c(1,2.5),
             c(1,2.5),
             c(1,2.5),
             c(1,2.5),
             c(1,2.5),
             c(1,NA))

gridExtra::grid.arrange(tree, pp, layout_matrix = lay)

```

or just with primary markers

```{r fig.width= 10}
# some more genes as landmarks
controls =list("genes related to L5/6"=c("Foxp2","Tbr1"), "genes related to L2/3"=c("Mef2c"), "genes related to Prog"=c("Nes","Sox2") , "genes related to L1"=c() , "genes related to L4"=c()) 
dend %>%
dendextend::set("labels", ifelse(labels(dend) %in% rownames(pca_1)[rownames(pca_1) %in% c(unlist(layers),unlist(controls))], labels(dend), "")) %>%
  dendextend::set("branches_k_color", value = c("gray80","#4DBBD5FF","#91D1C2FF" ,"gray80","#F39B7FFF","#E64B35FF","#3C5488FF"), k = 7) %>%
plot(horiz=F, axes=T,ylim = c(0,100))

```




Now we can plot the PCA

```{r}
# dataframe to be able to label only primary markers and control genes
textdf <- pca_1[rownames(pca_1) %in% c(unlist(layers),unlist(controls)) , ]

 for (m in c(1:length(controls))) {
  for (g in controls[[m]]) {
    if(g %in% rownames(textdf)){
      textdf[g,"highlight"] = names(controls[m])
    } 
  }
}

# deciding the colors
mycolours <- c("genes related to L5/6" = "#3C5488FF","genes related to L2/3"="#F39B7FFF","genes related to Prog"="#4DBBD5FF","genes related to L1"="#E64B35FF","genes related to L4" = "#91D1C2FF", "not marked"="#B09C85FF")

# to assing correcly the cluster number and the color
mycolours2 = c("Reln","Satb2","Rorb","Bcl11b","Vim")
names(mycolours2) = unique(cut[unlist(layers)])

mycolours2[mycolours2 == "Reln"] = "#E64B35FF"
mycolours2[mycolours2 == "Satb2"] = "#F39B7FFF"
mycolours2[mycolours2 == "Rorb"] = "#91D1C2FF"
mycolours2[mycolours2 == "Bcl11b"] = "#3C5488FF"
mycolours2[mycolours2 == "Vim"] = "#4DBBD5FF"
color_to_add = unique(pca_1$hclust)[!unique(pca_1$hclust) %in% as.numeric(names(mycolours2))]
names(color_to_add) = unique(pca_1$hclust)[!unique(pca_1$hclust) %in% as.numeric(names(mycolours2))]
color_to_add[color_to_add %in% 
                 unique(pca_1$hclust)[!unique(pca_1$hclust) %in% as.numeric(names(mycolours2))]] = "#B09C85FF"
mycolours2 = c(mycolours2,color_to_add)

pca1 = ggplot(subset(pca_1,!hclust %in% unique(cut[unlist(layers)])  ), aes(x=PC1, y=PC2)) +  geom_point(alpha = 0.3,color = "#B09C85FF",size=1)

#pca2 = pca1 + geom_point(data = subset(pca_1, highlight != "not marked" ), aes(x=PC1, y=PC2, colour=hclust),size=2.5,alpha = 0.9) 
pca2 = pca1 + geom_point(data = subset(pca_1, hclust %in% unique(cut[unlist(layers)]) ), aes(x=PC1, y=PC2, colour=as.character(hclust)),size=1,alpha = 0.5) + 
 scale_color_manual( "Status", values = mycolours2)  +
  scale_fill_manual( "Status", values = mycolours2)  +
  xlab("") + ylab("") +
  geom_label_repel(data =textdf , aes(x = PC1, y = PC2, label = rownames(textdf),fill=as.character(hclust)),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction = "both",
                   na.rm=TRUE,
                   seed = 1234) +
  geom_label_repel(data =textdf , aes(x = PC1, y = PC2, label = rownames(textdf)),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 1, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234) +
  ggtitle("PCA") +
  theme_light(base_size=10) +
    theme(axis.text.x=element_blank(),plot.title = element_text(size=14, 
                                    face="italic", 
                                    color="#3C5488FF",
                                    hjust=0.01,
                                    lineheight=1.2,margin = margin(t = 5, b = -15)),
        axis.text.y=element_blank(),
        legend.position = "none")  # titl)

pca2 #+ geom_encircle(data = pca_1, aes(group=hclust)) 
```

t-SNE code and plot

```{r}
# run the t-SNE
cl.genes.tsne = Rtsne(g.space ,initial_dims = 100, dims = 2, perplexity=30,eta = 200, verbose=F, max_iter = 3000,theta=0.4,num_threads = 10,pca_center = T, pca_scale = FALSE, normalize = T )

d_tsne_1 = as.data.frame(cl.genes.tsne$Y)
rownames(d_tsne_1) = rownames(g.space)

d_tsne_1 = d_tsne_1[order.dendrogram(dend),]

# save the cluster numebr inside a dataframe with the t-SNE information
d_tsne_1$hclust = cut

d_tsne_1$names = rownames(d_tsne_1)

# as before to label only some genes
textdf <- d_tsne_1[rownames(d_tsne_1) %in% c(unlist(layers),unlist(controls)),]

for (m in c(1:length(controls))) {
  for (g in controls[[m]]) {
    if(g %in% rownames(textdf)){
      textdf[g,"highlight"] = names(controls[m])
    } 
  }
}


 p1 = ggplot(subset(d_tsne_1,!hclust %in% unique(cut[unlist(layers)])), aes(x=V1, y=V2)) +  geom_point(alpha = 0.3, color = "#B09C85FF", size=1)

p2 = p1 + geom_point(data = subset(d_tsne_1, hclust %in% unique(cut[unlist(layers)]) ), aes(x=V1, y=V2, colour=as.character(hclust)),size=1,alpha = 0.5) +
    scale_color_manual("Status", values = mycolours2)  +
  scale_fill_manual("Status", values = mycolours2)  +
  xlab("") + ylab("")+
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = names,fill=as.character(hclust)),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction = "both",
                   na.rm=TRUE,
                   seed = 1234) +
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = names),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 1, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234) +
  ggtitle("t-SNE") +
  theme_light(base_size=10) +
  theme(axis.text.x=element_blank(),plot.title = element_text(size=14, 
                                    face="italic", 
                                    color="#3C5488FF",
                                    hjust=0.01,
                                    lineheight=1.2,margin = margin(t = 5, b = -15)),
        axis.text.y=element_blank(),
        legend.position = "none")  # titl)
p2
```
Code to create an iteractive plot. This can be modified to be used with all the plots.

```{r echo=TRUE}

p = ggplot(d_tsne_1, aes(x=V1, y=V2, text= paste("gene: ",names))) +  
  geom_point(size=2, aes(colour=as.character(hclust)), alpha=0.8) +
  scale_color_manual("Status", values = mycolours2)  +
  xlab("") + ylab("") +
  ggtitle("t-SNE") +
  theme_light(base_size=10) +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank())

ggplotly(p)
```

Multidimensional scaling (MDS) and plot
```{r}
# run the MDS
genes.dist.euc =  dist(g.space, method =  "euclidean")
#fit <- isoMDS(genes.dist.euc) # not linear
fit <- isoMDS(genes.dist.euc)

fit.genes = as.data.frame(fit$points)

fit.genes = fit.genes[order.dendrogram(dend),]

fit.genes$hclust = cut


fit.genes$names = rownames(fit.genes)

mycolours3 <- c("cluster L5/6 markers" = "#3C5488FF","cluster L2/3 markers"="#F39B7FFF","cluster Prog markers"="#4DBBD5FF","cluster L1 markers"="#E64B35FF","cluster L4 markers" = "#91D1C2FF", "not identified cluster"="#B09C85FF")

#mycolours3 <- c("cluster layer V-VI markers" = "#3C5488FF","cluster layer II-III markers"="#F39B7FFF","cluster progenitor markers"="#4DBBD5FF","cluster layer I markers"="#E64B35FF","cluster layer IV markers" = "#91D1C2FF", "not identified cluster"="#B09C85FF")


#fit.genes$hclust = factor(cutree(hc.norm, 7))
used = vector()
for (k in c(1:length(layers))) {
  #print(k)
  tt =as.numeric(cut[layers[[k]]][1])
  fit.genes[fit.genes$hclust == tt,"cluster"] = paste("cluster",names(layers[k]),"markers", sep = " " )
  used = c(used,cut[layers[[k]]][1])
}

fit.genes[fit.genes$hclust %in% (unique(fit.genes$hclust)[!unique(fit.genes$hclust) %in% used]),]$cluster = "not identified cluster"

textdf <- fit.genes[rownames(fit.genes) %in% c(unlist(layers),unlist(controls)),]

   f1 = ggplot(subset(fit.genes,!hclust %in% unique(cut[unlist(layers)]) ), aes(x=V1, y=V2)) +  geom_point(alpha = 0.3, color = "#B09C85FF", size=1)

f2 = f1 + geom_point(data = subset(fit.genes, hclust %in% unique(cut[unlist(layers)]) ), 
                     aes(x=V1, y=V2, colour=cluster), size=1,alpha = 0.5) +
  scale_color_manual("Status", values = mycolours3,
                     labels = c("Layer I cluster ","Layers II/III cluster",
                                "Layer IV cluster",
                                "Layers V/VI cluster",
                                "Progenitors cluster") )  +
  scale_fill_manual("Status", values = mycolours3)  + 
  xlab("") + ylab("")+
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = rownames(textdf),fill=cluster),
                   label.size = NA, 
                   alpha = 0.5, 
                   direction ="both",
                   na.rm=TRUE,
                   seed = 1234, show.legend = FALSE) +
  geom_label_repel(data =textdf , aes(x = V1, y = V2, label = rownames(textdf)),
                   label.size = NA, 
                   segment.color = 'black',
                   segment.size = 0.5,
                   direction = "both",
                   alpha = 1, 
                   na.rm=TRUE,
                   fill = NA,
                   seed = 1234, show.legend = FALSE) +
  ggtitle("MDS") +
  theme_light(base_size=10) +
  theme(axis.text.x=element_blank(),plot.title = element_text(size=14, 
                                    face="italic", 
                                    color="#3C5488FF",
                                    hjust=0.01,
                                    lineheight=1.2,margin = margin(t = 5, b = -15)),
        axis.text.y=element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(color = "#3C5488FF",face ="italic" ),
        legend.position = "bottom")  # titl)


f2 + scale_y_reverse()+ scale_x_reverse()#+ geom_encircle(data = fit.genes, aes(group=`5_clusters`)) 
```
```{r eval=FALSE, include=FALSE}
library("AnnotationDbi")
library("org.Mm.eg.db")
library("GO.db")

```
```{r eval=FALSE, include=FALSE}
GO = "GO:0003676" #nucleic acid binding
library(org.Mm.eg.db)
list = select(org.Mm.eg.db, keys = rownames(pca_1),
              columns=c("SYMBOL","GOALL"),keytype="SYMBOL") #[fit.genes$hclust %in% c(7,4,1,2,3),]),
list = list[list$GOALL == GO,]
list = list[complete.cases(list),]
not.useful = c("NAS","TAS","IEA","IC","ND")
list1 = list[!list$EVIDENCEALL %in% not.useful,]
list1 = unique(list1$SYMBOL)
pca_1$GO = NA
pca_1[list1,]$GO = "nucleic acid binding"
pca_1[list1,c(11:ncol(pca_1))]
```
```{r}
sessionInfo()
```

