---
title: "Loo et all markers"
output:
  html_notebook:
    collapsed: no
    css: html-md-01.css
    fig_caption: yes
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r}
library(COTAN)
library(ggplot2)
library(data.table)
library(Matrix)
library(factoextra)
library(ggrepel)
library(dendextend)
```


# Gene clustering using COTAN

```{r}
input_dir = "Data/"
layers = list("L1"=c("Reln","Lhx5"), "L2/3"=c("Satb2","Cux1"), "L4"=c("Rorb","Sox5") , "L5/6"=c("Bcl11b","Fezf2") , "Prog"=c("Vim","Hes1"))
objE17 = readRDS(file = "Data/E17_cortex_cl2.cotan.RDS")
```

```{r}
g.space = get.gene.coexpression.space(objE17, 
                                      n.genes.for.marker = 25,  primary.markers = unlist(layers))
```
```{r}
g.space = as.data.frame(as.matrix(g.space))

coex.pca.genes <- prcomp(t(g.space),
                 center = TRUE,
                 scale. = F) 

fviz_eig(coex.pca.genes, addlabels=TRUE,ncp = 10)
#fviz_eig(coex.pca.genes, choice = "eigenvalue", addlabels=TRUE)
```

## Hierarchical clustering

```{r fig.width=10}
controls =list("genes related to L5/6"=c("Foxp2","Tbr1"), "genes related to L2/3"=c("Mef2c"), "genes related to Prog"=c("Nes","Sox2") , "genes related to L1"=c() , "genes related to L4"=c())
# clustering usign Ward method
hc.norm = hclust(dist(g.space), method = "ward.D2")

# and cut the tree into 5 clusters (for example)
cut = cutree(hc.norm, k = 7, order_clusters_as_data = F)

# It crates the tree
dend <- as.dendrogram(hc.norm)

# I can use a dataframe from the pca to store some data regarding the clustering
pca_1 = as.data.frame(coex.pca.genes$rotation[,1:5])
pca_1 = pca_1[order.dendrogram(dend),]


mycolours <- c("genes related to L5/6" = "#3C5488FF","genes related to L2/3"="#F39B7FFF","genes related to Prog"="#4DBBD5FF","genes related to L1"="#E64B35FF","genes related to L4" = "#91D1C2FF", "not marked"="#B09C85FF")

# save the cluster number into the dataframe
pca_1$hclust = cut

dend =branches_color(dend,k=7,col=c("#4DBBD5FF","#91D1C2FF","#F39B7FFF","#E64B35FF","#3C5488FF","#91D1C2FF","#B09C85FF" ),groupLabels = T)
#dend =color_labels(dend,k=5,labels = rownames(pca_1),col=pca_1$colors)



dend %>%
set("labels", ifelse(labels(dend) %in% rownames(pca_1)[rownames(pca_1) %in% c(unlist(layers),unlist(controls))], labels(dend), "")) %>%
  set("branches_k_color", value = c("gray80","#4DBBD5FF","#91D1C2FF" ,"gray80","#F39B7FFF","#E64B35FF","#3C5488FF"), k = 7) %>%
plot(horiz=F, axes=T,ylim = c(0,100))
```

## Import of markers from Loo et al. paper

```{r}
Markers_Loo = read.csv("Data/Markers_Loo.csv")

genes_detected = pca_1[rownames(pca_1) %in% unlist(Markers_Loo)[!unlist(Markers_Loo) == ""],]

```

Gene present in the 10% of most differentially expressed genes by COTAN
```{r}
dim(genes_detected)[1]
```

of total genes detected as markers by Loo et al
```{r}
length(unlist(Markers_Loo)[!unlist(Markers_Loo) == ""])
```

Number of genes detected:
```{r}
sum(unlist(Markers_Loo)[!unlist(Markers_Loo) == ""] %in% rownames(objE17@coex))
```
Removed becouse not detected
```{r}
pure.markers = unlist(Markers_Loo)
pure.markers = pure.markers[!unlist(Markers_Loo) == ""]
pure.markers[!pure.markers %in% rownames(objE17@coex)]
```
```{r}
Markers_Loo = as.list(Markers_Loo)
for (i in names(Markers_Loo)) {
    Markers_Loo[[i]] = Markers_Loo[[i]][! Markers_Loo[[i]] == ""]
}
Markers_Loo
```
Primary markers also used by Loo et al.
```{r}
unlist(layers)[unlist(layers) %in% unique(unlist(Markers_Loo))]
```



Table

```{r}
tableMarkersCOTAN = as.data.frame(matrix(nrow = 6,ncol = 4))
colnames(tableMarkersCOTAN)=c("Loo.L.I","Loo.L.II.IV","Loo.L.V.VI","Loo.PROG")
rownames(tableMarkersCOTAN)=c("COTAN.L.I","COTAN.L.II.III","COTAN.L.IV","COTAN.L.V.VI","COTAN.PROG","COTAN.Not Grouped")
tableMarkersCOTAN_no_sec = tableMarkersCOTAN

layers.cluster = c("Reln","Satb2","Rorb","Bcl11b","Vim")
names(layers.cluster) = unique(cut[unlist(layers)])
layers.cluster
```

## Comparition between Loo et al. markers and COTAN markers

```{r}
groups = list("L.I"=names(which(layers.cluster == "Reln"))
              ,"L.II.III"=names(which(layers.cluster == "Satb2")),
              "L.IV"=names(which(layers.cluster == "Rorb")),
              "L.V.VI"=names(which(layers.cluster == "Bcl11b")),
              "PROG"=names(which(layers.cluster == "Vim")))

for(layer1 in c("L.I","L.II.III", "L.IV","L.V.VI","PROG")){
    for(layer2 in c("L.I","L.II.IV","L.V.VI","PROG")){
    tableMarkersCOTAN[paste0("COTAN.",layer1),paste0("Loo.",layer2)] =
        sum(rownames(pca_1[pca_1$hclust %in% groups[[layer1]],]) %in% Markers_Loo[[layer2]])
    tableMarkersCOTAN[paste0("COTAN.","Not Grouped"),paste0("Loo.",layer2)] =
        sum(rownames(pca_1[!pca_1$hclust %in% unlist(groups),]) %in% Markers_Loo[[layer2]])
    }
}

tableMarkersCOTAN
```
Without secondary markers

```{r}
pca_2 = pca_1[! rownames(pca_1) %in% colnames(g.space),]


for(layer1 in c("L.I","L.II.III", "L.IV","L.V.VI","PROG")){
    for(layer2 in c("L.I","L.II.IV","L.V.VI","PROG")){
    tableMarkersCOTAN_no_sec[paste0("COTAN.",layer1),paste0("Loo.",layer2)] =
        sum(rownames(pca_2[pca_2$hclust %in% groups[[layer1]],]) %in% Markers_Loo[[layer2]])
    tableMarkersCOTAN_no_sec[paste0("COTAN.","Not Grouped"),paste0("Loo.",layer2)] =
        sum(rownames(pca_2[!pca_2$hclust %in% unlist(groups),]) %in% Markers_Loo[[layer2]])
    }
}

tableMarkersCOTAN_no_sec
```


```{r}
specific.genes.table = data.frame("genes"=c(), "COTAN"=c(),"Loo."=c())
tt1 = c()
tt2 = c()
for(layer1 in c("L.I","L.II.III", "L.IV","L.V.VI","PROG")){
    for(layer2 in c("L.I","L.II.IV","L.V.VI","PROG")){
    tt1 = data.frame("genes"= rownames(pca_1[pca_1$hclust %in% groups[[layer1]],])[rownames(pca_1[pca_1$hclust 
                          %in% groups[[layer1]],]) %in% Markers_Loo[[layer2]]])
    if (dim(tt1)[1] > 0) {
                   tt1 = cbind(tt1,  "COTAN"=layer1, "Loo."=layer2)
        }
    tt2 = data.frame("genes"= 
        rownames(pca_1[!pca_1$hclust %in% unlist(groups),])[rownames(pca_1[!pca_1$hclust %in%                   unlist(groups),]) %in% Markers_Loo[[layer2]]])
    if (dim(tt2)[1] > 0) {
        tt2 = cbind(tt2, "COTAN"= "Not Grouped", "Loo."=layer2)
    }
    specific.genes.table = rbind(specific.genes.table,tt1,tt2)
    }
}

specific.genes.table[!(duplicated(specific.genes.table)) , ]

```
```{r fig.height= 80, fig.width= 8}
dend %>%
set("labels", ifelse(labels(dend) %in% rownames(pca_1)[rownames(pca_1) %in% specific.genes.table$genes], labels(dend), "")) %>%
  set("branches_k_color", value = c("gray80","#4DBBD5FF","#91D1C2FF" ,"gray80","#F39B7FFF","#E64B35FF","#3C5488FF"), k = 7) %>%
    plot(horiz=T, axes=T,xlim = c(0,80),cex = 10, dLeaf = -7)

```

# WGCNA 

```{r}
library(WGCNA)
library(cluster)
library(data.table)
library(Matrix)
library(Seurat)
library(utils)
library(dplyr)
library(patchwork)
library(graphics)
options(stringsAsFactors = FALSE)
data_dir = "Data/"
library(gplots)
myheatcol = colorpanel(250,'red',"orange",'lemonchiffon')
```


## Test with the 2000 most varied genes

This seems the best option for the analysis with WGCNA.

```{r}
data = as.data.frame(fread(paste(input_dir,"E175_only_cortical_cells.txt.gz", sep = "/"),sep = "\t"))
data = as.data.frame(data)
rownames(data) = data$V1
data = data[,2:ncol(data)]
data[1:10,1:10]
```

```{r}
E17 <- CreateSeuratObject(counts = data, project = "Cortex E17.5", min.cells = 3, min.features = 200)
E17[["percent.mt"]] <- PercentageFeatureSet(E17, pattern = "^mt-")
VlnPlot(E17, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
hist(E17$nFeature_RNA/E17$nCount_RNA, breaks = 100)
```
```{r}
plot1 <- FeatureScatter(E17, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(E17, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```
```{r}
E17 <- NormalizeData(E17, normalization.method = "LogNormalize", scale.factor = 10000)
E17 <- FindVariableFeatures(E17, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(E17), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(E17)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```
```{r}
all.genes <- rownames(E17)
E17 <- ScaleData(E17, features = all.genes)
E17 <- RunPCA(E17, features = VariableFeatures(object = E17))
DimPlot(E17, reduction = "pca")
```


```{r}
all.genes <- rownames(E17)
E17 <- ScaleData(E17, features = all.genes)

seurat.data = as.matrix(E17[["RNA"]]@data)

Var.genes = VariableFeatures(E17)[1:2000]

```


```{r}
datExpr0 = t(seurat.data[rownames(seurat.data) %in% Var.genes,])
gsg = goodSamplesGenes(datExpr0, verbose = 3)
gsg$allOK
if (!gsg$allOK){
    # Optionally, print the gene and sample names that were removed:
    if (sum(!gsg$goodGenes)>0)
        printFlush(paste("Removing genes:", paste(names(datExpr0)[!gsg$goodGenes], collapse = ", ")));
    if (sum(!gsg$goodSamples)>0)
        printFlush(paste("Removing samples:", paste(rownames(datExpr0)[!gsg$goodSamples], collapse = ", ")));
    # Remove the offending genes and samples from the data:
    datExpr0 = datExpr0[gsg$goodSamples, gsg$goodGenes]
}



sampleTree = hclust(dist(datExpr0), method = "average");
# Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# The user should change the dimensions if the window is too large or too small.
sizeGrWindow(12,9)

par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
     cex.axis = 1.5, cex.main = 2)

# Plot a line to show the cut
#abline(h = 400, col = "red")
```
No outliner detected

```{r}
# Automatic network construction and module detection
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 10, to=25, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(datExpr0, powerVector = powers, verbose = 5)
```
Plot the results:
```{r}
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
Tested with 5, 3 and 2 and 4. The best seems 2
```{r}
net = blockwiseModules(datExpr0, power = 2, maxBlockSize = 20000,
                       TOMType = "signed", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE,
                       saveTOMFileBase = "E17.5",
                       verbose = 3)

```
```{r}
# open a graphics window
sizeGrWindow(12, 9)
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

```
## Comparition between Loo et al. markers and WGCNA markers

```{r}
plotNetworkHeatmap(datExpr0, plotGenes = unlist(Markers_Loo),
networkType="signed", useTOM=TRUE,
power=2, main="D. TOM in an signed network")

```
```{r}
plotNetworkHeatmap(datExpr0, plotGenes = unlist(Markers_Loo),
networkType="unsigned", useTOM=TRUE,
power=2, main="D. TOM in an unsigned network")

```

```{r}
plotNetworkHeatmap(datExpr0, plotGenes = unique(unlist(layers)),
networkType="signed", useTOM=TRUE,
power=2, main="D. TOM in an signed network")

```
```{r}
plotNetworkHeatmap(datExpr0, plotGenes = unique(unlist(layers)),
networkType="unsigned", useTOM=TRUE,
power=2, main="D. TOM in an unsigned network")

```

```{r}
# Calculate topological overlap anew: this could be done more efficiently by saving the TOM
# calculated during module detection, but let us do it again here.
dissTOM = 1-TOMsimilarityFromExpr(datExpr0, power = 2);
# Transform dissTOM with a power to make moderately strong connections more visible in the heatmap
plotTOM = dissTOM^7;
# Set diagonal to NA for a nicer plot
diag(plotTOM) = NA;
rownames(dissTOM)=colnames(datExpr0)
colnames(dissTOM)=colnames(datExpr0)

selectTOM = dissTOM[rownames(dissTOM) %in% unlist(Markers_Loo),colnames(dissTOM) %in% unlist(Markers_Loo)];
# There’s no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster.
selectTree = hclust(as.dist(selectTOM), method = "average")
moduleColors = mergedColors
names(moduleColors) = rownames(dissTOM)
selectColors = moduleColors[names(moduleColors) %in% unlist(Markers_Loo)]
# Open a graphical window
sizeGrWindow(9,9)
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing
# the color palette; setting the diagonal to NA also improves the clarity of the plot
plotDiss = selectTOM^7;
diag(plotDiss) = NA;
TOMplot(plotDiss, selectTree, selectColors, main = "Network heatmap plot, selected genes",col=myheatcol)

```

```{r}
plotDendroAndColors(selectTree,selectColors,
                     "Module colors",
                     dendroLabels = NULL, hang = 0.03,
                     addGuide = TRUE, guideHang = 0.05)
```

```{r}
net$colors[names(net$colors) %in% unlist(Markers_Loo)]
```

WGCNA using the 2000 genes most varied by Seurat, detects the following number of markers.

```{r}
length(net$colors[names(net$colors) %in% unlist(Markers_Loo)])
```

```{r}
tableMarkersWGCNA = as.data.frame(matrix(nrow = 5,ncol = 4))
colnames(tableMarkersWGCNA)=c("Loo.L.I","Loo.L.II.IV","Loo.L.V.VI","Loo.PROG")
rownames(tableMarkersWGCNA)=c("WGCNA.L.I","WGCNA.Not Grouped","WGCNA.PROG","WGCNA.unknown1", "WGCNA.unknown2")

net$colors[unique(unlist(layers))]
```


```{r}
# Attention! The next list need to be updated by hand!
groups = list("L.I"=4,"unknown1"=3,"PROG"=1, "unknown2"=2 , "Not Grouped"=0 )

for(layer1 in c("L.I","unknown1","PROG","unknown2","Not Grouped")){
    for(layer2 in c("L.I","L.II.IV", "L.V.VI","PROG")){
    tableMarkersWGCNA[paste0("WGCNA.",layer1),paste0("Loo.",layer2)] =
        sum(names(net$colors[net$colors %in% groups[[layer1]]]) %in% Markers_Loo[[layer2]])
    #tableMarkersWGCNA[paste0("WGCNA.","Not Grouped"),paste0("Loo.",layer2)] =
     #   sum(names(net$colors[!net$colors %in% unlist(groups)]) %in% Markers_Loo[[layer2]])
    }
}

tableMarkersWGCNA
```




```{r fig.width= 10, fig.height= 50}

dend %>%
set("labels", ifelse(labels(dend) %in% rownames(pca_1)[rownames(pca_1) %in% unlist(Markers_Loo)], labels(dend), "")) %>%
  set("branches_k_color", value = c("gray80","#4DBBD5FF","#91D1C2FF" ,"gray80","#F39B7FFF","#E64B35FF","#3C5488FF"), k = 7) %>%
    set("labels_cex"=0.1) %>%
    set("branches_lwd", 0.5) %>%
plot_horiz.dendrogram(horiz=T, axes=T,xlim = c(0,80), dLeaf = -15,text_pos = 3)

```


```{r}
sessionInfo()
```







