---
title: "WGCNA experiments"
output:
    html_notebook:
      css: html-md-01.css
      fig_caption: yes
      highlight: haddock
      number_sections: yes
      theme: spacelab
      toc: yes
      toc_float: true
      collapsed: no
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r}
workingDir = "."
library(WGCNA)
library(cluster)
library(data.table)
library(Matrix)
library(Seurat)
library(utils)
library(dplyr)
library(patchwork)
library(graphics)
options(stringsAsFactors = FALSE)
data_dir = "Data/"
library(gplots)
myheatcol = colorpanel(250,'red',"orange",'lemonchiffon')

```

# Seurat analysis and normalization

```{r eval=FALSE, include=FALSE}
download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/samples/GSM2861nnn/GSM2861514/suppl/GSM2861514_E175_Only_Cortical_Cells_DGE.txt.gz",
              "Data/E175_only_cortical_cells.txt.gz",method = "wget", quiet = F)
```
```{r}
data = as.data.frame(fread(paste(data_dir,"E175_only_cortical_cells.txt.gz", sep = "/"),sep = "\t"))
data = as.data.frame(data)
rownames(data) = data$V1
data = data[,2:ncol(data)]
data[1:10,1:10]
```

```{r}
E17 <- CreateSeuratObject(counts = data, project = "Cortex E17.5", min.cells = 3, min.features = 200)
E17[["percent.mt"]] <- PercentageFeatureSet(E17, pattern = "^mt-")
VlnPlot(E17, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
```{r}
hist(E17$nFeature_RNA/E17$nCount_RNA, breaks = 100)
```
```{r}
plot1 <- FeatureScatter(E17, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(E17, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```
```{r}
E17 <- NormalizeData(E17, normalization.method = "LogNormalize", scale.factor = 10000)
E17 <- FindVariableFeatures(E17, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(E17), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(E17)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```
```{r}
all.genes <- rownames(E17)
E17 <- ScaleData(E17, features = all.genes)
E17 <- RunPCA(E17, features = VariableFeatures(object = E17))
DimPlot(E17, reduction = "pca")
```

# WGCNA

## Test usign all genes normalized by Seurat

```{r}
all.genes <- rownames(E17)
E17 <- ScaleData(E17, features = all.genes)

seurat.data = as.matrix(E17[["RNA"]]@data)

datExpr0 = t(seurat.data)
gsg = goodSamplesGenes(datExpr0, verbose = 3)
gsg$allOK
if (!gsg$allOK){
    # Optionally, print the gene and sample names that were removed:
    if (sum(!gsg$goodGenes)>0)
        printFlush(paste("Removing genes:", paste(names(datExpr0)[!gsg$goodGenes], collapse = ", ")));
    if (sum(!gsg$goodSamples)>0)
        printFlush(paste("Removing samples:", paste(rownames(datExpr0)[!gsg$goodSamples], collapse = ", ")));
    # Remove the offending genes and samples from the data:
    datExpr0 = datExpr0[gsg$goodSamples, gsg$goodGenes]
}


sampleTree = hclust(dist(datExpr0), method = "average")
# Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# The user should change the dimensions if the window is too large or too small.
sizeGrWindow(12,9)
#pdf(file = "Plots/sampleClustering.pdf", width = 12, height = 9);
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
     cex.axis = 1.5, cex.main = 2)

```
No outlier detected.

Automatic network construction and module detection
Choose a set of soft-thresholding powers
```{r}
powers = c(c(1:10), seq(from = 10, to=20, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(datExpr0, powerVector = powers, verbose = 5)
# Plot the results:
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
Thresholds tested: 2, 3, 6
The best is 2.
```{r}
net = blockwiseModules(datExpr0, power = 2, maxBlockSize = 20000,
                       TOMType = "signed", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE,
                       saveTOMFileBase = "E17.5",
                       verbose = 3)

# open a graphics window
sizeGrWindow(12, 9)
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
```
```{r}
primary.markers = c("Reln","Lhx5","Cux1","Satb2","Rorb","Sox5","Fezf2","Bcl11b","Vim","Hes1")

net$colors[primary.markers]
```

As we can see the primary markers are not well distributed: it detect layer I and progenitors but all other genes are in the same cluster (module).

Plot the dendrogram and the module colors underneath
```{r}
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)
```



```{r}
plotNetworkHeatmap(datExpr0, plotGenes = primary.markers,
networkType="unsigned", useTOM=TRUE,
power=2, main="D. TOM in an unsigned network")
```


```{r}
plotNetworkHeatmap(datExpr0, plotGenes = primary.markers,
networkType="signed", useTOM=TRUE,
power=2, main="D. TOM in an signed network")
```

```{r}
gene.sets.list = list("primary.markers"=primary.markers,
                      "NPGs"=c("Nes","Vim","Sox2","Sox1","Notch1", "Hes1","Hes5","Pax6"), 
                      "RG" = c("Vim","Sox2","Pax6","Hes5","Hes1"),
                      "IN" = c("Tubb3","Tbr1","Stmn1","Neurod1","Dcx"),
                      "PNGs"=c("Map2","Tubb3","Neurod1","Nefm","Nefl","Dcx","Tbr1"))#,
                      #"constitutive" =c("Calm1","Cox6b1","Ppia","Rpl18","Cox7c", "Erh","H3f3a","Taf1b","Taf2","Gapdh","Actb", "Golph3", "Mtmr12", "Zfr", "Sub1", "Tars", "Amacr") )

genes = unique(unlist(gene.sets.list))

plotNetworkHeatmap(datExpr0, plotGenes = genes, networkType="signed", useTOM=TRUE, 
                   power=2, main="D. TOM in an signed network")

```


```{r}
net$colors[primary.markers]
```
```{r}
net$colors[unique(unlist(gene.sets.list))]
```
```{r}
# Calculate topological overlap anew: this could be done more efficiently by saving the TOM
# calculated during module detection, but let us do it again here.
dissTOM = 1-TOMsimilarityFromExpr(datExpr0, power = 2);
# Transform dissTOM with a power to make moderately strong connections more visible in the heatmap
plotTOM = dissTOM^7;
# Set diagonal to NA for a nicer plot
diag(plotTOM) = NA;
rownames(dissTOM)=colnames(datExpr0)
colnames(dissTOM)=colnames(datExpr0)

selectTOM = dissTOM[primary.markers, primary.markers]
# There’s no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster.
selectTree = hclust(as.dist(selectTOM), method = "average")

moduleColors = mergedColors
names(moduleColors) = rownames(dissTOM)
selectColors = moduleColors[primary.markers]
# Open a graphical window
sizeGrWindow(9,9)
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing
# the color palette; setting the diagonal to NA also improves the clarity of the plot
plotDiss = selectTOM^7;
diag(plotDiss) = NA;
TOMplot(plotDiss, selectTree, selectColors, main = "Network heatmap plot, selected genes",col=myheatcol)

```

```{r}
plotDendroAndColors(selectTree,selectColors,
                     "Module colors",
                     dendroLabels = NULL, hang = 0.03,
                     addGuide = TRUE, guideHang = 0.05)
```


## Test with the 2000 most varied genes

```{r}
Var.genes = VariableFeatures(E17)[1:2000]

```


```{r}
datExpr0 = t(seurat.data[rownames(seurat.data) %in% Var.genes,])
gsg = goodSamplesGenes(datExpr0, verbose = 3)
gsg$allOK
if (!gsg$allOK){
    # Optionally, print the gene and sample names that were removed:
    if (sum(!gsg$goodGenes)>0)
        printFlush(paste("Removing genes:", paste(names(datExpr0)[!gsg$goodGenes], collapse = ", ")));
    if (sum(!gsg$goodSamples)>0)
        printFlush(paste("Removing samples:", paste(rownames(datExpr0)[!gsg$goodSamples], collapse = ", ")));
    # Remove the offending genes and samples from the data:
    datExpr0 = datExpr0[gsg$goodSamples, gsg$goodGenes]
}



sampleTree = hclust(dist(datExpr0), method = "average");
# Plot the sample tree: Open a graphic output window of size 12 by 9 inches
# The user should change the dimensions if the window is too large or too small.
sizeGrWindow(12,9)
#pdf(file = "Plots/sampleClustering.pdf", width = 12, height = 9);
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5,
     cex.axis = 1.5, cex.main = 2)

# Plot a line to show the cut
#abline(h = 400, col = "red")
```
No outliner detected

```{r}
# Automatic network construction and module detection
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 10, to=25, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(datExpr0, powerVector = powers, verbose = 5)
```
Plot the results:
```{r}
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
```
Tested with 5, 3 and 2 and 4. The best seems 2
```{r}
net = blockwiseModules(datExpr0, power = 2, maxBlockSize = 20000,
                       TOMType = "signed", minModuleSize = 30,
                       reassignThreshold = 0, mergeCutHeight = 0.25,
                       numericLabels = TRUE, pamRespectsDendro = FALSE,
                       saveTOMs = TRUE,
                       saveTOMFileBase = "E17.5",
                       verbose = 3)

```
```{r}
# open a graphics window
sizeGrWindow(12, 9)
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(net$dendrograms[[1]], mergedColors[net$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05)

```
```{r}
plotNetworkHeatmap(datExpr0, plotGenes = unique(unlist(gene.sets.list)),
networkType="signed", useTOM=TRUE,
power=2, main="D. TOM in an signed network")

```
```{r}
plotNetworkHeatmap(datExpr0, plotGenes = primary.markers,
networkType="signed", useTOM=TRUE,
power=2, main="D. TOM in an signed network")

```
```{r}
# Calculate topological overlap anew: this could be done more efficiently by saving the TOM
# calculated during module detection, but let us do it again here.
dissTOM = 1-TOMsimilarityFromExpr(datExpr0, power = 2);
# Transform dissTOM with a power to make moderately strong connections more visible in the heatmap
plotTOM = dissTOM^7;
# Set diagonal to NA for a nicer plot
diag(plotTOM) = NA;
rownames(dissTOM)=colnames(datExpr0)
colnames(dissTOM)=colnames(datExpr0)

selectTOM = dissTOM[primary.markers, primary.markers];
# There’s no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster.
selectTree = hclust(as.dist(selectTOM), method = "average")
moduleColors = mergedColors
names(moduleColors) = rownames(dissTOM)
selectColors = moduleColors[primary.markers]
# Open a graphical window
sizeGrWindow(9,9)
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing
# the color palette; setting the diagonal to NA also improves the clarity of the plot
plotDiss = selectTOM^7;
diag(plotDiss) = NA;
TOMplot(plotDiss, selectTree, selectColors, main = "Network heatmap plot, selected genes",col=myheatcol)

```

```{r}
plotDendroAndColors(selectTree,selectColors,
                     "Module colors",
                     dendroLabels = NULL, hang = 0.03,
                     addGuide = TRUE, guideHang = 0.05)
```

## Comparition with Loo et al. markers

```{r}
net$colors[primary.markers]
```
```{r}
net$colors[genes]
```
```{r}
Markers_Loo = read.csv("Data/Markers_Loo.csv")

Markers_Loo
```

```{r}
tableMarkersWGCNA = as.data.frame(matrix(nrow = 4,ncol = 4))
colnames(tableMarkersWGCNA)=c("Loo.L.I","Loo.L.II.IV","Loo.L.V.VI","Loo.PROG")
rownames(tableMarkersWGCNA)=c("WGCNA.L.I","WGCNA.L.II.VI","WGCNA.PROG","WGCNA.Not Groupped")

net$colors[primary.markers]
```


```{r}
groups = list("L.I"=4,"L.II.VI"=0,"PROG"=1)

for(layer1 in c("L.I","L.II.VI","PROG")){
    for(layer2 in c("L.I","L.II.IV", "L.V.VI","PROG")){
    tableMarkersWGCNA[paste0("WGCNA.",layer1),paste0("Loo.",layer2)] =
        sum(names(net$colors[net$colors %in% groups[[layer1]]]) %in% Markers_Loo[[layer2]])
    tableMarkersWGCNA[paste0("WGCNA.","Not Groupped"),paste0("Loo.",layer2)] =
        sum(names(net$colors[!net$colors %in% unlist(groups)]) %in% Markers_Loo[[layer2]])
    }
}

tableMarkersWGCNA
```
```{r}
selectTOM = dissTOM[rownames(dissTOM) %in% unlist(Markers_Loo),colnames(dissTOM) %in%  unlist(Markers_Loo)]
# There’s no simple way of restricting a clustering tree to a subset of genes, so we must re-cluster.
selectTree = hclust(as.dist(selectTOM), method = "average")
moduleColors = mergedColors
names(moduleColors) = rownames(dissTOM)
selectColors = moduleColors[names(moduleColors) %in% unlist(Markers_Loo)]
# Open a graphical window
sizeGrWindow(9,9)
# Taking the dissimilarity to a power, say 10, makes the plot more informative by effectively changing
# the color palette; setting the diagonal to NA also improves the clarity of the plot
plotDiss = selectTOM^7;
diag(plotDiss) = NA;
TOMplot(plotDiss, selectTree, selectColors, main = "Network heatmap plot, markers from Loo et al.",col=myheatcol)
```
```{r}
plotDendroAndColors(selectTree,selectColors,
                     "Module colors",
                     dendroLabels = NULL, hang = 0.03,
                     addGuide = TRUE, guideHang = 0.05)
```


```{r}
sessionInfo()
```

